\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}
\usepackage{xurl}
\usepackage{microtype}
\usepackage{enumitem}

\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  citecolor=blue,
  urlcolor=blue
}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{remark}{Remark}
\theoremstyle{definition}
\newtheorem{definition}{Definition}

\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\R}{\mathbb{R}}

\title{Preliminary Notes on ERM with Few Latent Tasks}
\author{}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
This note records preliminary findings for the following question. Data are generated hierarchically: first draw latent task variables
\[
\theta_1,\dots,\theta_k \overset{\mathrm{iid}}{\sim} p(\theta),
\]
then for each task draw
\[
Z_{ij} = (X_{ij},Y_{ij}) \overset{\mathrm{iid}}{\sim} P_{\theta_i},
\qquad j=1,\dots,m.
\]
We pool all $km$ examples and run empirical risk minimization (ERM) over a hypothesis class $\mathcal{H}$. When does the resulting predictor achieve low risk under the mixture distribution $D(z) = \int P_{\theta}(z)\,p(d\theta)$, especially in regimes where the number of distinct tasks $k$ is much smaller than the usual $O(\varepsilon^{-2})$ scale? The literature suggests that this problem sits between classical supervised learning, meta-learning, and domain generalization, but is not identical to any of them. The main preliminary conclusion is simple: pooled ERM is unbiased for the target mixture risk, but the real bottleneck is task-level heterogeneity. Large $m$ only helps with within-task noise; beating the $k$ bottleneck requires structural assumptions on the task-risk map $\theta \mapsto \E[\ell(h;Z)\mid \theta]$.
\end{abstract}

\section{Problem setup}
Let $\ell(h;z)$ be a loss in $[0,1]$. The population risk of a predictor $h \in \mathcal{H}$ is
\[
R(h) := \E_{\theta \sim p}\E_{Z \sim P_{\theta}}[\ell(h;Z)].
\]
Given hierarchical data $\{Z_{ij}\}_{i \in [k], j \in [m]}$, the pooled empirical risk is
\[
\widehat{R}_{k,m}(h) := \frac{1}{k}\sum_{i=1}^k \frac{1}{m}\sum_{j=1}^m \ell(h;Z_{ij}),
\]
and pooled ERM returns
\[
\widehat{h} \in \arg\min_{h \in \mathcal{H}} \widehat{R}_{k,m}(h).
\]

Two features distinguish this from ordinary iid learning.
\begin{enumerate}[label=(\roman*)]
\item The individual examples $Z_{ij}$ are not jointly iid because points from the same task share the same latent $\theta_i$.
\item The dependence is highly structured: the sample is a collection of $k$ iid clusters, each of size $m$.
\end{enumerate}

The question of interest is not whether $km$ examples are enough in the classical iid sense. Rather, it is whether a small number of distinct tasks can still suffice because the task family has special structure.

\section{Literature review}

\subsection{Motivation from in-context learning}
The motivating analogy is the line of work showing that in-context learning can recover learning algorithms that generalize beyond the finite task family seen during pretraining. Akyurek et al.~\cite{akyurek2023} showed that transformers can implement linear estimators such as gradient descent and ridge regression in context. Raventos et al.~\cite{raventos2023} sharpened the point by training transformers on a finite set of linear regression tasks and identifying a task-diversity threshold: below it, the transformer behaves like a Bayesian estimator tied to the pretraining task prior; above it, behavior aligns with ridge regression over a broader task family.

The present supervised-learning question is a natural analogue. Instead of asking whether a transformer trained on finitely many tasks can perform the right in-context algorithm on new tasks, we ask whether ordinary pooled ERM, trained on finitely many latent tasks, can recover a predictor with low risk under the full meta-distribution.

\subsection{Learning to learn from an environment of tasks}
Baxter's classical ``environment of tasks'' framework~\cite{baxter2000} is the closest foundational theory. The learner observes multiple tasks sampled from a task environment and uses them to learn an inductive bias that transfers to novel tasks. The central sample-complexity split is exactly the one relevant here: number of tasks versus number of examples per task.

Later work made this split more explicit. Maurer, Pontil, and Romera-Paredes~\cite{maurer2016} analyzed multitask and learning-to-learn representation learning, identifying regimes where many related tasks reduce the burden on per-task sample size. Denevi et al.~\cite{denevi2019} studied learning-to-learn with biased regularization and showed benefits when the variance across tasks is small. More recently, Aliakbarpour et al.~\cite{aliakbarpour2024} proved that in structured representation-learning settings, one can metalearn from very few samples per task, but only because the shared representation class is strongly restricted.

These papers do not analyze pooled ERM directly. Typically the algorithm learns a shared representation, bias vector, or initialization and then adapts to a fresh task. Still, they strongly suggest the right principle for the present problem: if one wants nontrivial generalization from few tasks, the family of task-level predictors or risks must be simple.

\subsection{Domain generalization and latent domains}
Another adjacent literature is domain generalization. Here multiple environments are available at training time and the goal is to perform well on unseen environments. IRM~\cite{arjovsky2019} and REx~\cite{krueger2021} search for predictors whose behavior is stable across environments. However, these methods usually assume observed environment labels and often target a stronger notion of robustness than the average-risk objective above.

For the current question, two facts from the domain-generalization literature matter. First, plain ERM is a strong baseline: Gulrajani and Lopez-Paz~\cite{gulrajani2020} showed that carefully tuned ERM is competitive with specialized domain-generalization methods on standard benchmarks, and Sener and Koltun~\cite{sener2023} argued that some domain-generalization penalties help only when they do not induce excess empirical risk. Second, there is work on latent domains, such as Matsuura and Harada~\cite{matsuura2019}, where domain labels are not observed and one must infer hidden environments.

That said, the objective here is still different. We do not seek worst-domain robustness, and we do not necessarily use domain labels at either train or test time. The target is the average mixture risk under $p(\theta)P_{\theta}$, with a clustered but exchangeable training sample.

\section{Preliminary findings}

\subsection{A brief tutorial on fat-shattering dimension}
The fat-shattering dimension is the real-valued analogue of VC dimension. If a function class is binary-valued, shattering means realizing every $0/1$ labeling of a finite set. For real-valued classes, one needs a scale parameter $\gamma > 0$: the class should realize every binary pattern with a margin of size $\gamma$ around some thresholds.

\begin{definition}[Fat-shattering dimension]
Let $\mathcal{F} \subseteq [0,1]^{\mathcal{X}}$ and let $\gamma > 0$. A set $x_1,\dots,x_t \in \mathcal{X}$ is \emph{$\gamma$-shattered} by $\mathcal{F}$ if there exist thresholds $s_1,\dots,s_t \in \R$ such that for every subset $A \subseteq [t]$, there exists $f_A \in \mathcal{F}$ satisfying
\[
f_A(x_i) \le s_i - \gamma
\quad \text{for } i \in A,
\qquad
f_A(x_i) \ge s_i + \gamma
\quad \text{for } i \notin A.
\]
The \emph{fat-shattering dimension} $\mathrm{fat}_{\gamma}(\mathcal{F})$ is the largest $t$ for which such a set exists, or $\infty$ if there is no largest such $t$.
\end{definition}

The learning problem characterized by the fat-shattering dimension is the following generic one. We have an example space $\mathcal{Z}$, a hypothesis class $\mathcal{H}$, and a bounded loss function
\[
\ell : \mathcal{H} \times \mathcal{Z} \to [0,1].
\]
An unknown distribution $P$ on $\mathcal{Z}$ is fixed, the learner receives iid samples
\[
z_1,\dots,z_n \overset{\mathrm{iid}}{\sim} P,
\]
and it wants to find a hypothesis whose population loss is nearly minimal:
\[
\E_{z \sim P}[\ell(\widehat{h},z)]
\le
\inf_{h \in \mathcal{H}} \E_{z \sim P}[\ell(h,z)] + \varepsilon.
\]
A natural algorithm is ERM:
\[
\widehat{h}
\in
\arg\min_{h \in \mathcal{H}} \frac{1}{n}\sum_{i=1}^n \ell(h,z_i).
\]
To connect this with fat-shattering, one passes to the induced loss class
\[
\mathcal{F}_{\ell}
:=
\{z \mapsto \ell(h,z) : h \in \mathcal{H}\}
\subseteq [0,1]^{\mathcal{Z}}.
\]
The fat-shattering dimension characterizes when this ERM problem is distribution-free learnable, meaning that for every $\varepsilon,\delta > 0$ there is an $n(\varepsilon,\delta)$ such that the above excess-risk guarantee holds with probability at least $1-\delta$, uniformly over all distributions $P$.

\begin{theorem}[Sample complexity via fat-shattering]
\label{thm:generic-fat-sample}
There exist universal constants $c,C > 0$ such that the following holds. Let $\mathcal{H}$ be a hypothesis class, let $\ell : \mathcal{H} \times \mathcal{Z} \to [0,1]$ be a bounded loss, and let
\[
\mathcal{F}_{\ell}
:=
\{z \mapsto \ell(h,z) : h \in \mathcal{H}\}
\subseteq [0,1]^{\mathcal{Z}}
\]
be the induced loss class. Fix any distribution $P$ on $\mathcal{Z}$ and draw iid samples
\[
z_1,\dots,z_n \overset{\mathrm{iid}}{\sim} P.
\]
If
\[
n
\ge
C \cdot \frac{\mathrm{fat}_{c\varepsilon}(\mathcal{F}_{\ell}) + \ln(1/\delta)}{\varepsilon^2},
\]
then with probability at least $1-\delta$, every empirical risk minimizer
\[
\widehat{h}
\in
\arg\min_{h \in \mathcal{H}} \frac{1}{n}\sum_{i=1}^n \ell(h,z_i)
\]
satisfies
\[
\E_{z \sim P}[\ell(\widehat{h},z)]
\le
\inf_{h \in \mathcal{H}} \E_{z \sim P}[\ell(h,z)] + \varepsilon.
\]
\end{theorem}

This theorem is the clean answer to where the scale parameter $\gamma$ goes. To learn to accuracy $\varepsilon$, one needs to evaluate the complexity of the loss class at scale $\gamma \asymp \varepsilon$, which is why the bound involves $\mathrm{fat}_{c\varepsilon}$ rather than a scale-free quantity.

\begin{theorem}[What fat-shattering characterizes for proper ERM]
\label{thm:generic-fat-erm}
Assume the usual measurability conditions. Let $\mathcal{H}$ be a hypothesis class, let $\ell : \mathcal{H} \times \mathcal{Z} \to [0,1]$ be a bounded loss, and let
\[
\mathcal{F}_{\ell}
:=
\{z \mapsto \ell(h,z) : h \in \mathcal{H}\}
\subseteq [0,1]^{\mathcal{Z}}.
\]
Then the following are equivalent:
\begin{enumerate}[label=(\roman*)]
\item \textbf{Distribution-free learnability by proper ERM.} For every $\varepsilon,\delta > 0$, there exists $n(\varepsilon,\delta)$ such that for every distribution $P$ on $\mathcal{Z}$, if $z_1,\dots,z_n \overset{\mathrm{iid}}{\sim} P$ and
\[
\widehat{h}
\in
\arg\min_{h \in \mathcal{H}} \frac{1}{n}\sum_{i=1}^n \ell(h,z_i),
\]
then with probability at least $1-\delta$,
\[
\E_{z \sim P}[\ell(\widehat{h},z)]
\le
\inf_{h \in \mathcal{H}} \E_{z \sim P}[\ell(h,z)] + \varepsilon.
\]

\item \textbf{Finite fat-shattering dimension at every scale.} For every $\gamma > 0$,
\[
\mathrm{fat}_{\gamma}(\mathcal{F}_{\ell}) < \infty.
\]
\end{enumerate}
\end{theorem}

This is the classical qualitative characterization for proper learning with bounded real-valued losses; see Bartlett, Long, and Williamson~\cite{bartlett1996}.

\begin{remark}[Scope of the characterization]
\label{rem:generic-fat-caution}
Theorem~\ref{thm:generic-fat-erm} is intentionally a theorem about \emph{proper ERM}. It says that fat-shattering is the right scale-sensitive notion for ERM and uniform-convergence arguments. It does \emph{not} say that the upper bound in Theorem~\ref{thm:generic-fat-sample} is minimax optimal among all distribution-free learners, especially improper ones. In fact, more recent work explicitly distinguishes ERM learnability from more general learnability notions in real-valued prediction problems; see Attias et al.~\cite{attias2024}.
\end{remark}

What one can safely conclude from Theorem~\ref{thm:generic-fat-sample} and Theorem~\ref{thm:generic-fat-erm} is the following. If $\mathrm{fat}_{\gamma}(\mathcal{F}_{\ell})$ stays bounded as $\gamma \downarrow 0$, then the standard ERM sample-complexity upper bound is of order $\varepsilon^{-2}$. If it grows like $\log(1/\gamma)$, the ERM bound picks up only a logarithmic overhead. If it grows like $\gamma^{-p}$, then the ERM bound scales like $\varepsilon^{-(p+2)}$. And if it grows exponentially in $1/\gamma$, then the standard ERM upper bound becomes exponential in $1/\varepsilon$.

Three facts are useful to keep in mind.
\begin{enumerate}[label=(\roman*)]
\item If $\mathcal{F}$ is $\{0,1\}$-valued and $\gamma < 1/2$, then $\mathrm{fat}_{\gamma}(\mathcal{F})$ reduces to the ordinary VC dimension.
\item The scale parameter matters. Large $\gamma$ asks for robust separation, while small $\gamma$ probes finer distinctions.
\item For bounded real-valued ERM, fat-shattering is the right complexity notion: it gives a qualitative iff characterization for proper ERM learnability and quantitative ERM sample-complexity upper bounds; see Theorems~\ref{thm:generic-fat-sample} and~\ref{thm:generic-fat-erm} and~\cite{bartlett1996,bartlettlong1998,colomboni2023}. The lower-bound story for arbitrary improper learners is subtler; see Remark~\ref{rem:generic-fat-caution} and Attias et al.~\cite{attias2024}.
\end{enumerate}

\subsection{A task-complexity characterization}
We now return to the hierarchical setting. After reducing to the idealized regime $m=\infty$, each hypothesis $h$ induces a real-valued function on the task space $\Theta$, namely its conditional risk profile $\theta \mapsto r_{\theta}(h)$.

With the fat-shattering dimension in hand, one can now define a genuine notion of task complexity. The correct object is a scale-sensitive complexity profile of the \emph{task-risk functions} induced by $\mathcal{H}$.

For each $\theta$, define the conditional risk profile
\[
r_{\theta}(h) := \E[\ell(h;Z)\mid \theta].
\]
The right way to package this is to view each $h$ as a real-valued function on the task space:
\[
g_h(\theta) := r_{\theta}(h),
\qquad
\mathcal{G}_{\mathcal{H}} := \{g_h : h \in \mathcal{H}\} \subseteq [0,1]^{\Theta}.
\]

\begin{definition}[Task complexity profile]
For $\gamma > 0$, define the \emph{task complexity profile} of $\mathcal{H}$ at scale $\gamma$ by
\[
\mathrm{TC}_{\mathcal{H}}(\gamma)
:=
\mathrm{fat}_{\gamma}(\mathcal{G}_{\mathcal{H}}),
\]
the fat-shattering dimension of the task-risk class $\mathcal{G}_{\mathcal{H}}$ at margin $\gamma$.

Equivalently, $\mathrm{TC}_{\mathcal{H}}(\gamma) \ge t$ if there exist tasks $\theta_1,\dots,\theta_t$ and thresholds $s_1,\dots,s_t$ such that for every subset $A \subseteq [t]$, there is some $h_A \in \mathcal{H}$ satisfying
\[
g_{h_A}(\theta_i) \le s_i - \gamma
\quad \text{for } i \in A,
\qquad
g_{h_A}(\theta_i) \ge s_i + \gamma
\quad \text{for } i \notin A.
\]
\end{definition}

The quantity $\mathrm{TC}_{\mathcal{H}}(\gamma)$ is a direct measure of task diversity \emph{as seen by the hypothesis class}. It asks how many tasks can be margin-separated in all possible patterns by choosing different hypotheses.

\begin{lemma}[Idealized reduction to iid ERM over tasks]
\label{lem:task-reduction}
Assume the idealized regime $m=\infty$, so that for each sampled task $\theta_i$ the learner has access to the exact task risks $r_{\theta_i}(h)$ for all $h \in \mathcal{H}$. Then pooled ERM returns
\[
\widehat{h}_k \in \arg\min_{h \in \mathcal{H}} \frac{1}{k}\sum_{i=1}^k r_{\theta_i}(h)
=
\arg\min_{h \in \mathcal{H}} \frac{1}{k}\sum_{i=1}^k g_h(\theta_i).
\]
Moreover, for every $h \in \mathcal{H}$,
\[
\E_{\theta \sim p}[g_h(\theta)]
=
\E_{\theta \sim p}[r_{\theta}(h)]
=
\E_{\theta \sim p}\E_{Z \sim P_{\theta}}[\ell(h;Z)]
=
R(h).
\]
Therefore the idealized hierarchical excess-risk problem is exactly ordinary ERM for the bounded real-valued function class $\mathcal{G}_{\mathcal{H}}$ on the iid sample $\theta_1,\dots,\theta_k \sim p$, both at the empirical level and at the population-risk level.
\end{lemma}

\begin{proof}
When $m=\infty$, the within-task empirical average equals its conditional expectation, so the pooled empirical risk becomes
\[
\widehat{R}_{k,\infty}(h)
=
\frac{1}{k}\sum_{i=1}^k \E[\ell(h;Z)\mid \theta_i]
=
\frac{1}{k}\sum_{i=1}^k r_{\theta_i}(h)
=
\frac{1}{k}\sum_{i=1}^k g_h(\theta_i).
\]
Also,
\[
\E_{\theta \sim p}[g_h(\theta)]
=
\E_{\theta \sim p}\E[\ell(h;Z)\mid \theta]
=
R(h).
\]
Since the latent tasks are iid from $p$, the sample $(\theta_1,\dots,\theta_k)$ is iid and the claim follows.
\end{proof}

\begin{theorem}[Task-complexity characterization of pooled ERM]
\label{thm:task-complexity-iff}
Assume $\ell(h;z) \in [0,1]$ and the usual measurability conditions under which the standard fat-shattering characterization of bounded real-valued ERM applies. In the idealized regime $m=\infty$, the following are equivalent:
\begin{enumerate}[label=(\roman*)]
\item \textbf{Distribution-free learnability by pooled ERM.} For every $\varepsilon,\delta > 0$, there exists $k(\varepsilon,\delta)$ such that for every distribution $p(\theta)$, if $\theta_1,\dots,\theta_k \overset{\mathrm{iid}}{\sim} p$ and
\[
\widehat{h}_k \in \arg\min_{h \in \mathcal{H}} \frac{1}{k}\sum_{i=1}^k r_{\theta_i}(h),
\]
then with probability at least $1-\delta$,
\[
R(\widehat{h}_k) \le \inf_{h \in \mathcal{H}} R(h) + \varepsilon.
\]

\item \textbf{Finite task complexity at every scale.} For every $\gamma > 0$,
\[
\mathrm{TC}_{\mathcal{H}}(\gamma) < \infty.
\]
\end{enumerate}
\end{theorem}

\begin{proof}
By Lemma~\ref{lem:task-reduction}, the idealized pooled-ERM problem is exactly the problem of empirical risk minimization over the bounded real-valued class $\mathcal{G}_{\mathcal{H}}$ on iid samples from the domain $\Theta$.

The classical characterization of bounded real-valued ERM states that such a class is distribution-free learnable by ERM if and only if its fat-shattering dimension is finite at every scale; see Bartlett, Long, and Williamson~\cite{bartlett1996}. Applying that theorem to $\mathcal{G}_{\mathcal{H}}$ gives the equivalence between (i) and (ii).

For completeness, the easy direction is the implication from uniform convergence to ERM consistency. If with probability at least $1-\delta$ one has
\[
\sup_{h \in \mathcal{H}}
\left|
\frac{1}{k}\sum_{i=1}^k g_h(\theta_i) - \E_{\theta \sim p}[g_h(\theta)]
\right|
\le \frac{\varepsilon}{2},
\]
then for any population minimizer $h^* \in \arg\min_{h \in \mathcal{H}} R(h)$ and any ERM $\widehat{h}_k$,
\begin{align*}
R(\widehat{h}_k) - R(h^*)
&\le
\left(
R(\widehat{h}_k) - \frac{1}{k}\sum_{i=1}^k g_{\widehat{h}_k}(\theta_i)
\right)
+
\left(
\frac{1}{k}\sum_{i=1}^k g_{h^*}(\theta_i) - R(h^*)
\right) \\
&\le
2 \sup_{h \in \mathcal{H}}
\left|
\frac{1}{k}\sum_{i=1}^k g_h(\theta_i) - \E[g_h(\theta)]
\right|
\le \varepsilon.
\end{align*}
The converse direction is exactly the necessity statement in the cited fat-shattering characterization.
\end{proof}

\begin{lemma}[Finite-$m$ learnability via the meta-loss class]
\label{lem:finite-m-reduction}
For each $m \ge 1$, define the meta-sample
\[
S := (Z_1,\dots,Z_m) \in \mathcal{Z}^m,
\]
drawn by first sampling $\theta \sim p$ and then drawing $Z_1,\dots,Z_m \overset{\mathrm{iid}}{\sim} P_{\theta}$. Also define the meta-loss class
\[
G_{h,m}(S) := \frac{1}{m}\sum_{j=1}^m \ell(h;Z_j),
\qquad
\mathcal{G}_{\mathcal{H},m} := \{G_{h,m} : h \in \mathcal{H}\}.
\]
Assume that
\[
\mathrm{fat}_{\gamma}(\mathcal{G}_{\mathcal{H},m}) < \infty
\qquad \text{for every } \gamma > 0.
\]
Then the original hierarchical learning problem at this fixed $m$ is distribution-free learnable by pooled ERM: for every $\varepsilon,\delta > 0$, there exists $k(\varepsilon,\delta,m)$ such that for every choice of $p(\theta)$ and conditional laws $P_{\theta}$, if
\[
\widehat{h}_{k,m}
\in
\arg\min_{h \in \mathcal{H}} \widehat{R}_{k,m}(h)
\qquad\text{where}\qquad
\widehat{R}_{k,m}(h)
:=
\frac{1}{k}\sum_{i=1}^k \frac{1}{m}\sum_{j=1}^m \ell(h;Z_{ij}),
\]
then with probability at least $1-\delta$,
\[
R(\widehat{h}_{k,m})
\le
\inf_{h \in \mathcal{H}} R(h) + \varepsilon.
\]
\end{lemma}

\begin{proof}
Define the population risk for the finite-$m$ meta-problem by
\[
R_m(h)
:=
\E_{\theta \sim p}\E_{Z_1,\dots,Z_m \sim P_{\theta}}
\left[\frac{1}{m}\sum_{j=1}^m \ell(h;Z_j)\right].
\]
For each task $i$, the tuple
\[
S_i := (Z_{i1},\dots,Z_{im})
\]
is drawn independently from the same meta-distribution on $\mathcal{Z}^m$, and by definition
\[
\widehat{R}_{k,m}(h)
=
\frac{1}{k}\sum_{i=1}^k \frac{1}{m}\sum_{j=1}^m \ell(h;Z_{ij})
=
\frac{1}{k}\sum_{i=1}^k G_{h,m}(S_i).
\]
So pooled ERM is exactly ERM on the bounded real-valued class $\mathcal{G}_{\mathcal{H},m}$ using the iid meta-samples $S_1,\dots,S_k$.

Because $\mathrm{fat}_{\gamma}(\mathcal{G}_{\mathcal{H},m}) < \infty$ for every $\gamma > 0$, the standard fat-shattering characterization of bounded real-valued ERM implies that the meta-problem is distribution-free learnable. Equivalently, for every $\varepsilon,\delta > 0$, there exists $k(\varepsilon,\delta,m)$ such that with probability at least $1-\delta$,
\[
\sup_{h \in \mathcal{H}}
\left|
\widehat{R}_{k,m}(h) - R_m(h)
\right|
\le \frac{\varepsilon}{2}.
\]
In particular, on this event,
\[
\widehat{R}_{k,m}(h) \le R_m(h) + \frac{\varepsilon}{2}
\qquad \text{for every } h \in \mathcal{H}.
\]
Therefore, for any population minimizer $h_m^* \in \arg\min_{h \in \mathcal{H}} R_m(h)$ and any pooled ERM $\widehat{h}_{k,m}$,
\begin{align*}
R_m(\widehat{h}_{k,m}) - R_m(h_m^*)
&\le
\left(
R_m(\widehat{h}_{k,m}) - \widehat{R}_{k,m}(\widehat{h}_{k,m})
\right)
+
\left(
\widehat{R}_{k,m}(h_m^*) - R_m(h_m^*)
\right) \\
&\le
2 \sup_{h \in \mathcal{H}}
\left|
\widehat{R}_{k,m}(h) - R_m(h)
\right|
\le \varepsilon.
\end{align*}
Thus pooled ERM learns with respect to $R_m$.

Finally, for every $h \in \mathcal{H}$,
\[
R_m(h)
=
\E_{\theta \sim p}\E_{Z_1,\dots,Z_m \sim P_{\theta}}
\left[\frac{1}{m}\sum_{j=1}^m \ell(h;Z_j)\right]
=
\E_{\theta \sim p}\E_{Z \sim P_{\theta}}[\ell(h;Z)]
=
R(h).
\]
So an excess-risk guarantee with respect to $R_m$ is exactly an excess-risk guarantee with respect to the original hierarchical objective $R$. This proves the claim.
\end{proof}

\begin{theorem}[Sample complexity for pooled ERM]
\label{thm:sample-complexity-pooled}
There exist universal constants $c,C > 0$ such that the following holds.

\begin{enumerate}[label=(\roman*)]
\item For every finite $m \ge 1$, if
\[
k
\ge
C \cdot \frac{\mathrm{fat}_{c\varepsilon}(\mathcal{G}_{\mathcal{H},m}) + \ln(1/\delta)}{\varepsilon^2},
\]
then with probability at least $1-\delta$, every pooled ERM $\widehat{h}_{k,m}$ satisfies
\[
R(\widehat{h}_{k,m}) \le \inf_{h \in \mathcal{H}} R(h) + \varepsilon.
\]

\item In the idealized regime $m=\infty$, if
\[
k
\ge
C \cdot \frac{\mathrm{TC}_{\mathcal{H}}(c\varepsilon) + \ln(1/\delta)}{\varepsilon^2},
\]
then with probability at least $1-\delta$, every pooled ERM $\widehat{h}_{k}$ satisfies
\[
R(\widehat{h}_{k}) \le \inf_{h \in \mathcal{H}} R(h) + \varepsilon.
\]
\end{enumerate}
\end{theorem}

\begin{proof}
For the finite-$m$ case, Lemma~\ref{lem:finite-m-reduction} shows that pooled ERM is ordinary ERM on the bounded real-valued class $\mathcal{G}_{\mathcal{H},m}$ with iid meta-samples $S_1,\dots,S_k$.

Apply Theorem~\ref{thm:generic-fat-sample} with the example space $\mathcal{Z}^m$, the hypothesis class $\mathcal{H}$, and the bounded loss
\[
\ell_m(h,S) := G_{h,m}(S) = \frac{1}{m}\sum_{j=1}^m \ell(h;Z_j).
\]
Its induced loss class is exactly $\mathcal{G}_{\mathcal{H},m}$. Therefore Theorem~\ref{thm:generic-fat-sample} yields
\[
\E[G_{\widehat h_{k,m},m}(S)]
\le
\inf_{h \in \mathcal{H}} \E[G_{h,m}(S)] + \varepsilon.
\]
By Lemma~\ref{lem:finite-m-reduction}, $\E[G_{h,m}(S)] = R(h)$ for every $h$, so this is exactly the finite-$m$ claim.

The idealized $m=\infty$ case is the same argument, using Lemma~\ref{lem:task-reduction} and the identity
\[
\mathrm{TC}_{\mathcal{H}}(\gamma) = \mathrm{fat}_{\gamma}(\mathcal{G}_{\mathcal{H}}).
\]
Indeed, Theorem~\ref{thm:generic-fat-sample} gives
\[
\E[g_{\widehat h_k}(\theta)]
\le
\inf_{h \in \mathcal{H}} \E[g_h(\theta)] + \varepsilon,
\]
and Lemma~\ref{lem:task-reduction} identifies $\E[g_h(\theta)]$ with $R(h)$ for every $h$.
\end{proof}

\begin{remark}[What the characterization says]
Theorem~\ref{thm:task-complexity-iff} gives the exact answer in the noise-free task limit, while Lemma~\ref{lem:finite-m-reduction} shows that the finite-$m$ problem is the same question after replacing each latent task by a noisy length-$m$ snapshot. So there really is an if-and-only-if statement, but the right complexity is a \emph{profile}
\[
\gamma \mapsto \mathrm{TC}_{\mathcal{H}}(\gamma),
\]
not a single scalar such as a task diameter or an average pairwise distance.
\end{remark}

\begin{remark}[From characterization to rates]
Theorem~\ref{thm:task-complexity-iff} is qualitative: it characterizes when pooled ERM is possible in principle. To obtain quantitative bounds on the number of tasks $k$, one needs upper bounds on the size or growth of $\mathrm{TC}_{\mathcal{H}}(\gamma)$, or on localized variants of this complexity profile. To beat the generic $1/\varepsilon^2$ task barrier, one also needs a fast-rate condition at the task level, such as a Bernstein or central condition; see van Erven et al.~\cite{vanerven2015}.
\end{remark}

\section{Takeaways and candidate directions}
At this stage, the picture seems to be the following.

\begin{enumerate}[label=(\arabic*)]
\item \textbf{The right language is real-valued, not binary.} Once each hypothesis is viewed as a task-risk function $\theta \mapsto r_{\theta}(h)$, the relevant complexity notion is the fat-shattering dimension rather than VC dimension.
\item \textbf{There is an exact task-level iff statement.} In the idealized regime $m=\infty$, pooled ERM works if and only if the induced task-risk class $\mathcal{G}_{\mathcal{H}} = \{\theta \mapsto r_{\theta}(h)\}$ has finite fat-shattering dimension at every scale.
\item \textbf{The right complexity is a profile, not a scalar distance.} The quantity $\mathrm{TC}_{\mathcal{H}}(\gamma) = \mathrm{fat}_{\gamma}(\mathcal{G}_{\mathcal{H}})$ measures how richly hypotheses can separate many tasks simultaneously. This is stronger than any single pairwise notion of task discrepancy.
\item \textbf{Finite $m$ is the same question with noisy task snapshots.} The finite-$m$ model reduces to ERM on the meta-class $\mathcal{G}_{\mathcal{H},m}$, so the exact iff remains a fat-shattering statement after lifting tasks to length-$m$ samples.
\item \textbf{The remaining challenge is quantitative.} The conceptual characterization is now clean; the next problem is to get interpretable upper bounds on $\mathrm{TC}_{\mathcal{H}}(\gamma)$ in concrete models and then convert them into sharp bounds on the number of tasks $k$.
\end{enumerate}

The main open question for the next step is no longer whether an iff statement exists; it does. The harder question is:
\begin{quote}
\itshape
Which concrete structural assumptions on $\mathcal{G}_{\mathcal{H}}$ yield sharp and interpretable bounds on $\mathrm{TC}_{\mathcal{H}}(\gamma)$, and hence on $k$?
\end{quote}
Finding the right answer here would give a supervised-learning counterpart to the finite-task-diversity phenomena already observed in in-context learning.

\begin{thebibliography}{16}

\bibitem{akyurek2023}
Ekin Akyurek, Dale Schuurmans, Jacob Andreas, Tengyu Ma, and Denny Zhou.
\newblock What learning algorithm is in-context learning? Investigations with linear models.
\newblock In \emph{International Conference on Learning Representations}, 2023.
\newblock \url{https://arxiv.org/abs/2211.15661}

\bibitem{raventos2023}
Allan Raventos, Mansheej Paul, Feng Chen, and Surya Ganguli.
\newblock Pretraining task diversity and the emergence of non-Bayesian in-context learning for regression.
\newblock arXiv:2306.15063, 2023.
\newblock \url{https://arxiv.org/abs/2306.15063}

\bibitem{baxter2000}
Jonathan Baxter.
\newblock A model of inductive bias learning.
\newblock \emph{Journal of Artificial Intelligence Research}, 12:149--198, 2000.
\newblock \url{https://arxiv.org/abs/1106.0245}

\bibitem{bartlett1996}
Peter L. Bartlett, Philip M. Long, and Robert C. Williamson.
\newblock Fat-shattering and the learnability of real-valued functions.
\newblock \emph{Journal of Computer and System Sciences}, 52(3):434--452, 1996.
\newblock \url{https://doi.org/10.1006/jcss.1996.0033}

\bibitem{bartlettlong1998}
Peter L. Bartlett and Philip M. Long.
\newblock Prediction, learning, uniform convergence, and scale-sensitive dimensions.
\newblock \emph{Journal of Computer and System Sciences}, 56(2):174--190, 1998.
\newblock \url{https://doi.org/10.1006/jcss.1997.1557}

\bibitem{colomboni2023}
Roberto Colomboni, Emmanuel Esposito, and Andrea Paudice.
\newblock An improved uniform convergence bound with fat-shattering dimension.
\newblock arXiv:2307.06644, 2023.
\newblock \url{https://arxiv.org/abs/2307.06644}

\bibitem{maurer2016}
Andreas Maurer, Massimiliano Pontil, and Bernardino Romera-Paredes.
\newblock The benefit of multitask representation learning.
\newblock \emph{Journal of Machine Learning Research}, 17(81):1--32, 2016.
\newblock \url{https://jmlr.org/beta/papers/v17/15-242.html}

\bibitem{denevi2019}
Giulia Denevi, Carlo Ciliberto, Riccardo Grazzi, and Massimiliano Pontil.
\newblock Learning-to-learn stochastic gradient descent with biased regularization.
\newblock arXiv:1903.10399, 2019.
\newblock \url{https://arxiv.org/abs/1903.10399}

\bibitem{aliakbarpour2024}
Maryam Aliakbarpour, Konstantina Bairaktari, Gavin Brown, Adam Smith, Nathan Srebro, and Jonathan Ullman.
\newblock Metalearning with very few samples per task.
\newblock arXiv:2312.13978, 2024.
\newblock \url{https://arxiv.org/abs/2312.13978}

\bibitem{attias2024}
Idan Attias, Steve Hanneke, Alkis Kalavasis, Amin Karbasi, and Grigoris Velegkas.
\newblock Optimal learners for realizable regression: PAC learning and online learning.
\newblock arXiv:2307.03848, 2024.
\newblock \url{https://arxiv.org/abs/2307.03848}

\bibitem{arjovsky2019}
Martin Arjovsky, Leon Bottou, Ishaan Gulrajani, and David Lopez-Paz.
\newblock Invariant risk minimization.
\newblock arXiv:1907.02893, 2019.
\newblock \url{https://arxiv.org/abs/1907.02893}

\bibitem{krueger2021}
David Krueger, Ethan Caballero, J\"orn-Henrik Jacobsen, Amy Zhang, Jonathan Binas, Dinghuai Zhang, Bernhard Sch\"olkopf, and Aaron Courville.
\newblock Out-of-distribution generalization via risk extrapolation (REx).
\newblock In \emph{International Conference on Machine Learning}, 2021.
\newblock \url{https://arxiv.org/abs/2003.00688}

\bibitem{gulrajani2020}
Ishaan Gulrajani and David Lopez-Paz.
\newblock In search of lost domain generalization.
\newblock arXiv:2007.01434, 2020.
\newblock \url{https://arxiv.org/abs/2007.01434}

\bibitem{matsuura2019}
Toshihiko Matsuura and Tatsuya Harada.
\newblock Domain generalization using a mixture of multiple latent domains.
\newblock arXiv:1911.07661, 2019.
\newblock \url{https://arxiv.org/abs/1911.07661}

\bibitem{sener2023}
Ozan Sener and Vladlen Koltun.
\newblock Domain generalization without excess empirical risk.
\newblock arXiv:2308.15856, 2023.
\newblock \url{https://arxiv.org/abs/2308.15856}

\bibitem{vanerven2015}
Tim van Erven, Peter D. Gr\"unwald, Nishant A. Mehta, Mark D. Reid, and Robert C. Williamson.
\newblock Fast rates in statistical and online learning.
\newblock \emph{Journal of Machine Learning Research}, 16(54):1793--1861, 2015.
\newblock \url{https://www.jmlr.org/papers/v16/vanerven15a.html}

\end{thebibliography}

\end{document}
